---
title: "parse_spreadsheets"
author: "bryan"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

## Introduction

* Nutrient analysis is done on autoanalyzer and the results are returned as an untidy spreadsheet.
* Need to read the spreadsheet
* eliminate un-needed data (calibration stuff mostly)
    * these are mostly !is.na(name)
* correct readings by subtracting out the blanks
    * analysis done in batches with each batch (usually) bracketed by 2 blanks at the beginning and two at the end
    * the last batch only has blanks at the beginning.
    
## data sources
* spreadsheets received from Adam Pimienta and stored in here::here('data') directory.
* current data files:

```{r data_sources, echo = FALSE}
# get a list of the files
files <- list.files(here::here('data/raw'))
# remove temp files
files <- files[grep("~", files, invert = TRUE)]

# print list of files
print(files)
```

## data steps
* write two functions to process the data
* function "nut_data" reads a spreadsheet with the nutrient data
    * selects and renames the columns
    * filter out the calibration readings (type == "unknown")
    * parse name into "sample_date" and "sample_site"
    * add header info to data file as columns "run_name", "run_date", "operator"
    * process blanks (see below) and correct concentrations (um - blank_um)

* function "find_outs": is a subroutine for function "nut_data"
    * reads formatted data.frame from "nut_data" for a specified nutrient ("n" or "p")
    * selects only the data for "blanks"
    * tests for and removes outliers using boxplot (outliers are values outside 1.5 x the IQR)
    * calculates the means of successive (if they exist) blanks
    * returns a data frame with blanks and means(blanks) for each sample
    
* process the 4 files received from Stephen Shivers
* save results as .csv in folder "output"
    
```{r nut_data}
#' function to organize autoanalyzer nutrient data and correct concentrations for blanks
#' 
#' @param xlsx an untidy excel spreadsheet with the autoanalyzer data
nut_data <- function(xlsx) {

#read data & header info
f <- read_excel(xlsx, skip = 5)
h <- read_excel(xlsx, n_max = 3, col_names = FALSE)

# rename columns
names(f) <- tolower(names(f)) 
f <- rename(f, peak = "peak#", 
                n_raw_ht = "raw ht", n_cor_ht = "cor ht", n_um = um,
                p_raw_ht = "raw ht__1", p_cor_ht = "cor ht__1", p_um = um__1)

# filter type == "unknown"
f <- filter(f, type == "Unknown")

# parse name
y <- as.data.frame(str_split_fixed(f$name, " ", 2))
f$sample_date <- y[ , 1]
f$sample_site <- y[ , 2]

# create new id to parse identifier duplicates
f$id <-unlist(strsplit(f$identifier, "dup"))

# add header info to file
h <- read_excel(path = paste0(here::here('data/raw'), "/", files[i]), n_max = 3, col_names = FALSE)
f$run_name <- unlist(strsplit(as.character(h[1, 1]), ": "))[2]
f$run_date <- unlist(strsplit(as.character(h[2, 1]), ": "))[2]
f$operator <- unlist(strsplit(as.character(h[3, 1]), ": "))[2]

# select and reorder fields
f <- select(f, name, sample_date, sample_site, peak, position, identifier, id,
            n_raw_ht, n_cor_ht, n_um, p_raw_ht, p_cor_ht, p_um,
            run_name, run_date, operator)

# use find_outs to remove outliers and get means for blanks by id
bn <- find_outs(f, 'n')
bp <- find_outs(f, 'p')

# add blanks data to samples
s <- filter(f, sample_site != "milliQ") %>%
      left_join(bn) %>%
      left_join(bp)

# correct n_um and p_um
out <- mutate(s, n_um_cor = n_um - n_mean_b_um, p_um_cor = p_um - p_mean_b_um)

return(out)
}
```

```{r find_outs}
#' function to find and remove outliers from blanks
#' 
#' @param df a dataframe of values from function ??????
#' @param nutrient this will be 'n' or 'p'
find_outs <- function(df, nutrient = 'n') {
# filter, select and rename fields
b <- filter(df, sample_site == "milliQ") %>%
        select(id1 = identifier, b_um = paste0(nutrient, "_um"))

# use boxplot to find outliers.  outliers are values outside 1.5 * the IQR
outliers <- boxplot(b$b_um, plot = FALSE)$out

# remove outliers
witch <- which(b$b_um %in% outliers)
if(length(witch) > 0) b <- slice(b, -witch)

# get means of consecutive blanks
b1 <- data.frame(id1 = as.numeric(c(1, b$id1)), id2 = as.numeric(c(b$id1, nrow(df))),
                 b_um1 = c(NA, b$b_um), b_um2 = c(b$b_um, NA)) 

b1 <- mutate(b1, mean_b_um = rowMeans(select(b1, b_um1, b_um2), na.rm = TRUE))

# loop through rows and create output
out <- c()
for(i in 1:nrow(b1)){
if(b1$id2[i]-b1$id1[i] > 1) {
a <- data.frame(id = seq((b1$id1[i] + 1), (b1$id2[i])), 
                b_um1 = b1$b_um1[i], b_um2 = b1$b_um2[i], mean_b_um = b1$mean_b_um[i])
out <- rbind(out, a)
}}

# add nutrient to names and ouput
out <- mutate(out, id = as.character(id))
names(out)[-1] <- paste0(nutrient, '_', names(out)[-1])
return(out)
}
```

```{r process_data}
# process all files
for(i in 1:length(files)){
x <- nut_data(paste0(here::here('data/raw'), "/", files[i]))
write.csv(x, paste0(here::here('output'), '/', x$run_name[1], '.csv'), row.names = FALSE)
}
```

## data definitions for added or modified fields

field  | data definition
----------- | -----------------------------------------------------------------------------------------
sample_date  | parsed from the name field. If this is a sample this the first of the name field
sample_site  | parsed from the name field. If this is a sample this the second part of the name field
id | parsed from the "identifier" field. Some identifiers have a number + "dup"; "dup" is removed
run_name | added from the header
run_date | added from the header
operator | added from the header
p_b_um1 | value of first phosphorus blank in a group
p_b_um2 | value of second phosphorus blank in a group
p_mean_b_um | mean of p_b_um1 & p_b_um2 (or value if one is missing)
p_um_cor | corrected phosphorus concentration (p_um - p_mean_b_um)  
n_b_um1 | value of first nitrogen blank in a group
n_b_um2 | value of second nitrogen blank in a group
n_mean_b_um | mean of n_b_um1 & n_b_um2 (or value if one is missing)
n_um_cor | corrected nitrogen concentration (n_um - n_mean_b_um) 


