---
title: "parse_spreadsheets"
author: "bryan"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

## Introduction

* Nutrient analysis is done on autoanalyzer and the results are returned as an untidy spreadsheet.
* Need to read the spreadsheet
* eliminate un-needed data (calibration stuff mostly)
    * these are mostly !is.na(name)
* correct readings by subtracting out the blanks
    * analysis done in batches with each batch (usually) bracketed by 2 blanks at the beginning and two at the end
    * the last batch only has blanks at the beginning.
    
## data sources
* spreadsheets received from Adam Pimienta and stored in here::here('data') directory.
* current data files:

```{r data_sources, echo = FALSE}
# get a list of the files
files <- list.files(here::here('data/raw'))
# remove temp files
files <- files[grep("~", files, invert = TRUE)]

# print list of files
print(files)
```

## data steps
* write two functions to process the data
* function "nut_data" reads a spreadsheet with the nutrient data
    * selects and renames the columns
    * filter out the calibration readings (type == "unknown")
    * add header info to data file as columns "run_name", "run_date", "operator"
    * process blanks (see below) and correct concentrations (um - blank_um)

* function "find_outs": is a subroutine for function "nut_data"
    * reads formatted data.frame from "nut_data" for a specified nutrient ("n" or "p")
    * selects only the data for "blanks"
    * tests for and removes outliers using boxplot (outliers are values outside 1.5 x the IQR)
    * calculates the means of successive (if they exist) blanks
    * returns a data frame with blanks and means(blanks) for each sample
    
* process the 4 files received from Stephen Shivers
* save results as .csv in folder "output"
    
```{r nut_data}
# i <-1; xlsx <- paste0(here::here('data/raw'), "/", files[i])

#' function to organize autoanalyzer nutrient data and correct concentrations for blanks
#' 
#' @param xlsx path to an untidy excel spreadsheet with the autoanalyzer data
nut_data <- function(xlsx) {

#read data & header info
f <- read_excel(xlsx, skip = 5)
h <- read_excel(xlsx, n_max = 3, col_names = FALSE)

# select & rename columns
f <- f[1:11]
names(f) <- c("peak", "position", "identifier", "name", "type",
              "n_raw_ht", "n_cor_ht", "n_um",
              "p_raw_ht", "p_cor_ht", "p_um")

# filter type == "unknown" and change non-numeric values of n_um and p_um to NA
f <- filter(f, type == "Unknown") %>%
      mutate(n_um = as.numeric(n_um), p_um = as.numeric(p_um))

# add header info to file
f$run_name <- unlist(strsplit(as.character(h[1, 1]), ": "))[2]
f$run_date <- unlist(strsplit(as.character(h[2, 1]), ": "))[2]
f$operator <- unlist(strsplit(as.character(h[3, 1]), ": "))[2]

# use find_outs to remove outliers and get means for blanks by id
bn <- find_outs(f, 'n')
bp <- find_outs(f, 'p')

# add blanks data to samples
s <- slice(f, -grep("milliQ", f$name)) %>%
      left_join(bn) %>%
      left_join(bp)

# correct n_um and p_um & reorder fields
out <- mutate(s, n_um_cor = n_um - n_mean_b_um, p_um_cor = p_um - p_mean_b_um) %>%
        select(peak, name, position, identifier,
               n_raw_ht, n_cor_ht, n_um, n_um_cor, n_b_um1, n_b_um2, n_mean_b_um, 
               p_raw_ht, p_cor_ht, p_um, p_um_cor, p_b_um1, p_b_um2, p_mean_b_um,
               run_name, run_date, operator)

# select and reorder fields
f <- select(f, peak, name, position, identifier, 
            n_raw_ht, n_cor_ht, n_um, p_raw_ht, p_cor_ht, p_um,
            run_name, run_date, operator)


return(out)
}
```

```{r find_outs}
#' function to find and remove outliers from blanks
#' 
#' @param df a dataframe of values from function ??????
#' @param nutrient this will be 'n' or 'p'
find_outs <- function(df, nutrient = 'n') {
# filter, select and rename fields
b <- slice(df, grep("milliQ", df$name)) %>%
        select(id1 = peak, b_um = paste0(nutrient, "_um"))

# use boxplot to find outliers.  outliers are values outside 1.5 * the IQR
outliers <- boxplot(b$b_um, plot = FALSE)$out

# change outliers to NA
witch <- which(b$b_um %in% outliers)
b$b_um[witch] <- NA


# get means of consecutive blanks
b1 <- data.frame(id1 = c(1, b$id1), id2 = c(b$id1, max(b$id1)),
                 b_um1 = c(NA, b$b_um), b_um2 = c(b$b_um, NA)) 

b1 <- mutate(b1, mean_b_um = rowMeans(select(b1, b_um1, b_um2), na.rm = TRUE))

# loop through rows and create output
out <- c()
for(i in 1:nrow(b1)){
if(b1$id2[i]-b1$id1[i] > 1) {
a <- data.frame(peak = seq((b1$id1[i] + 1), (b1$id2[i])), 
                b_um1 = b1$b_um1[i], b_um2 = b1$b_um2[i], mean_b_um = b1$mean_b_um[i])
out <- rbind(out, a)
}}

# add nutrient to names and ouput
names(out)[-1] <- paste0(nutrient, '_', names(out)[-1])
return(out)
}
```

```{r process_data}
# process all files
for(i in 1:length(files)){
x <- tryCatch(nut_data(paste0(here::here('data/raw'), "/", files[i])), 
              error = function(e) 
                print(paste0("Warning: ", files[i], " did not parse")))
if(is.data.frame(x)) write.csv(x, paste0(here::here('output'), '/', x$run_name[1], '.csv'), 
                               row.names = FALSE)
}
```

## data definitions

field  | data definition
----------- | -----------------------------------------------------------------------------------------
peak | sequential order of samples - some calibration samples removed so not complete
name | the sample name often this is sample date and site
position | not sure what this is.  Probably indicates position sample was loaded into analyzer
identifier | appear to be a unique identifier for sample.  Some are labelled "dup" for duplicate.
n_raw_ht | from the operator.  Added "n_" to indicate this is a nitrogen sample	
n_cor_ht | from the operator.  Added "n_" to indicate this is a nitrogen sample	
n_um | from the operator.  Added "n_" to indicate this is a nitrogen sample.  This is the uncorrected nitrogen concentration.	
n_um_cor | corrected nitrogen concentration (n_um - n_mean_b_um); NOTE if NA this could mean that n_um &/or n_mean_b_um is missing.  Check manually.
n_b_um1 | value of first nitrogen blank in a group of observations; NA indicates value is an outlier or missing.
n_b_um2 | value of second nitrogen blank in a group of observations; NA indicates value is an outlier or missing.
n_mean_b_um | mean of the two blanks [(n_b_um1 + n_b_um2) / 2]; NA indicates both blanks are missing or outliers.
p_raw_ht | from the operator.  Added "p_" to indicate this is a phosphorus sample	
p_cor_ht | from the operator.  Added "p_" to indicate this is a phosphorus sample	
p_um | from the operator.  Added "p_" to indicate this is a phosphorus sample.  This is the uncorrected phosphorus concentration.	
p_um_cor | corrected phosphorus concentration (p_um - p_mean_b_um); NOTE if NA this could mean that p_um &/or p_mean_b_um is missing.  Check manually.
p_b_um1 | value of first phosphorus blank in a group of observations; NA indicates value is an outlier or are missing.
p_b_um2 | value of second phosphorus blank in a group of observations; NA indicates value is an outlier or missing.
p_mean_b_um | mean of the two blanks [(p_b_um1 + p_b_um2) / 2]; NA indicates both blanks are missing or are outliers.
run_name | added from the header; this must be the identifier for the sample run.
run_date | added from the header; this must be the date the sample was run.
operator | added from the header; this must be the person who ran the sample.


	



